\chapter{Discussion}
%1. Begin with a restatement of your research question, followed by a statement about whether or not, and how much, your findings "answer" the question.  These should be the first two pieces of information the reader encounters.
%ci sono delle mutazioni deleterie nel DNA mitocondriale o nei geni espressi nel mitocondrio?
The genetic causes of miscarriages are not yet fully known, there are areas of the genome that are still completely unexplored and  mtDNA is a significant element in this kind of study, in fact,through life mutations can occur in one mtDNA and its descendants creating genetic variability of mitochondrial DNA, some of these mutations might be deleterious. So, the analysis of the mtDNA sequences and the nuclear genes involved in mitochondrial processes, of embryos from pregnancy loss can informs on genes putatively causative of diseases. \\

%1a breve ricapitolazione dei risultati 
With this work I demonstrated that it is possible to identify variants in mitochondrial DNA to cause diseases using high-coverage whole-genome sequence data.
I analyzed the mtDNA sequences of the 10 embryos from mothers that experienced pregnancy loss to discover 276 unique variable sites in all samples. Per each sample, the number of variable sites vary between 22 and 103. Using a filtering procedure  based on genomic annotations I identified 4 variants that are likely to be deleterious. These variants are found in the \textit {MT-ND2}, \textit{MT-ATP6} and \textit{MT-ND6} genes. 
Pathogenic variants of the mitochondrial gene \textit{MT-ND2} are known to cause mtDNA-associated Leigh syndrome, as are variants of \textit{MT-ATP6} and \textit{MT-ND6}. 

In addition to variants of mtDNA I have analyzed variation in nuclear genes whose product are involved in mitochondrial processes. Three variants (rs2159132, rs10935321, rs16872235) seems to be relevant, either because they are shared among samples and because they are
the most deleterious according to vep. These variants are found in the \textit{COX10}, and \textit{MRPS22} \textit{GFM2} genes. These variants are associated with elements of the respiratory chain and mutations in these components are associated with mitochondrial hereditary disorders such as Leigh Syndrome.\\

The identification of heteroplasmic variants is an essential point in the analysis of the mitochondrial genome because very often deleterious variants are in heteroplasmy. My report identified 231 heteroplasmic sites. Among these, the percentage of heteroplasmy varies from a minimum of 0.000100  to a maximum of 0.999900. In particular, the embryo AS054 has the highest number of heteroplasmic sites and very high heteroplasmy (0.9998) at position 8668 (COSV62293167) in the \textit{MT-ATP6} gene responsible for Leigh syndrome.\\


Finally, I determined the haplogroup to  define the geographic distribuition of phylogenetic clusters among samples. My analysis reveals that eight out of eleven sequences belong to haplogroup H and T typical of West Eurasia, one sequence belong to haplogroup L typical of Africa, and two to haplogroups M and N which are commonly found in Asia. My findings about Haplogroups are concordant with the demographic data that were avaialble for these samples. \\


%2. Relate your findings to the issues you raised in the introduction. Note similarities, differences, common or different trends.  Show how your study either corraborates, extends, refines, or conflicts with previous findings.%3. If you have unexpected findings, try to interpret them in terms of method, interpretation, even a restructured hypothesis; in extreme cases, you may have to rewrite your introduction. Be honest about the limitations of your study.
% rispetto a quanto detto nel paragrafo dell'introduzione mt DNA e diseases (e sopratutto in miscarriages) che cosa aggiungi/confermi/discrediti

Most of the mitochondrial diseases are related to the respiratory chain and the processes that involve it. In according to what already available in literature the analysis I carried out confirm that there are mutations in the gene relevant for these processes. In addition, the variants that I found to be putatively deleterious are all associated with the same mitochondrial disorders, the Leigh syndrome. This suggests that these variants may actually be implicated in major disorders involving more severe phenotypic conditions.\\
 \\

%4. State the major conclusions from your study and present the theoretical and practical implications of your study.

Overall, I demonstrated that the analysis of mitochondrial DNA can help to clarify part of the causes of pregnancy loss and is informative about implications of deleterious variants of mitochondrial DNA that can lead to very serious disease. This study hopes to provides essential indications for the realization of a larger study.



%5. Discuss the implications of your study for future research and be specific about the next logical steps for future researchers.




